A Mechanised Library for Generic (Co)Programming

### 2024-02-13 paper/paper.tex
Containers and polynomial functors

However, there still remains the issue of reasoning about the equality of
container extensions. The contents of the container extensions need to be
accessed via a function \mintinline{Coq}{contents}. The domain of this function
depends on the shape of the container. This means that, for a given type of
shapes and positions, the equality of two objects of a container extension will
need to deal with heterogeneous equality. Since we want to remain axiom-free,
one way to deal with this is to rely on a \emph{decidable} validity predicate.
This way, we reach our final definition for container extensions. First,
containers are defined as pairs of shape and position types that have a
\emph{decidable} validity predicate. We represent decidable predicates as
functions that return a boolean, and lift booleans to propositions similarly to
SSReflect.
\begin{minted}{Coq}
Class Cont (Shape : Set) (Pos : Set) := { valid : Shape -> Pos -> bool }.
Coercion is_true (b : bool) : Prop := b = true.
\end{minted}
Then, we define \emph{container extensions} as
\begin{minted}{Coq}
Record App `(C : Cont Shape Pos) (X : Set) :=
  { shape : Shape; contents : {p | valid shape pos} -> X }.
\end{minted}

The functorial action for container extensions on morphisms boils down to
function composition, which trivially preserves identity and composition.
\begin{minted}{Coq}
Definition fmap `(C : Cont Shape Pos) `(f : A -> B) (x : App C A) : App C B
  := {| shape := shape x; contents := fun e => f (contents x e) |}
\end{minted}

@@

@@

\dcas{Examples of polynomials as container extensions?}

@@

\begin{minted}{Coq}
  Inductive AppR `{F : Container Sh P} (X : Type) {e : equiv X}
           (x y : App F X) : Prop :=
  | AppR_ext
      (Es : shape x =e shape y)
      (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 =e cont y e2).
\end{minted}
The reason for this is to avoid a circularity in the definitions: arbitrary
equivalences for positions would force us to define the contents as morphisms
from positions to the contained elements, but that would as well force us to
redefine container extensions so that they are only applicable to types with
arbitrary equalities and morphisms.  We will see in the following section why
this is problematic for defining initial algebras of container extensions.
\dcas{I am actually not sure of what I wrote above, and it may be something I
simply overlooked. I will look into the Coq code and try to generalise it if I
have the time}

@@

  Corollary cata_ana_hylo `(F : Container Sh P) `{equiv A} `{equiv B}
    (g : Alg F B) (h : RCoalg F A)
  : cata g \o rana h =e hylo g h.
\end{minted}

>>>>

However, there still remains the issue of reasoning about the equality of
container extensions. The contents of the container extensions need to be
accessed via a function \mintinline{Coq}{contents}. The domain of this function
depends on the shape of the container. This means that, for a given type of
shapes and positions, the equality of two objects of a container extension will
need to deal with heterogeneous equality. Since we want to remain axiom-free,
one way to deal with this is to rely on a \emph{decidable} validity predicate.
This way, we reach our final definition for container extensions. First,
containers are defined as pairs of shape and position types that have a
\emph{decidable} validity predicate. We represent decidable predicates as
functions that return a boolean, and lift booleans to propositions similarly to
SSReflect.
\begin{minted}{Coq}
Class Cont (Shape : Set) (Pos : Set) := { valid : Shape -> Pos -> bool }.
Coercion is_true (b : bool) : Prop := b = true.
\end{minted}
Then, we define \emph{container extensions} as
\begin{minted}{Coq}
Record App `(C : Cont Shape Pos) (X : Set) :=
  { shape : Shape; contents : {p | valid shape pos} -> X }.
\end{minted}
The functorial action for container extensions on morphisms boils down to
function composition, which trivially preserves identity and composition.
\begin{minted}{Coq}
Definition fmap `(C : Cont Shape Pos) `(f : A -> B) (x : App C A) : App C B
  := {| shape := shape x; contents := fun e => f (contents x e) |}
\end{minted}

@@

Polynomial functors can be defined as container extensions~\ref{fig:poly}. We
omit here the proofs that these container extensions are isomorphic to
polynomials, but this proof is straightforward. To illustrate how the identity,
constant, sum and product container extensions work, consider the
tree-generating functor (i.e. the functor whose fixpoint is a binary tree):
$F\; X = \mathbb{N} + X \times X$.

This functor is built as the sum of two functors: $X \mapsto \mathbb{N}$ and $X \mapsto
X \times X$. Consider first $X \mapsto \mathbb{N}$. This is a constant functor
that maps $X$ to the type $\mathbb{N}$. More generally, a functor that, given a
type $A$, maps everything to it, $X \mapsto A$, only needs one shape ($A$, L9
of Fig~\ref{fig:poly}), and zero positions (In Coq, \mintinline{Coq}{False},
L10 of Fig~\ref{fig:poly}). The reason why there are zero valid positions in
a constant container extension is because, given something of type 
\mintinline{Coq}{App (Const (Ks A) Kp) X}, it should be impossible to get
something of type \mintinline{Coq}{X}. But from the container shape, we will
always be able to get the element of type $A$\footnote{%
  Our decision to use \mintinline{Coq}{False} is arbitrary. In general, we
  could have chosen any time, with a validity predicate that always returns
  \mintinline{Coq}{false}. It would make no difference for code extraction.
}. This functor is, therefore \mintinline{Coq}{Const nat}.

Consider now $X \mapsto X \times X$. This is the product of two
identity functors $X \mapsto X$. The identity functor only has one shape, and
we can use the unit type to represent such shapes. From this shape we should be
able to get an element of type $X$, and we can capture this by using the unit
type for positions. To build an element of the container extension type, we
will need to provide a function of type \mintinline{Coq}{unit -> X}, which is
isomorphic to \mintinline{Coq}{X}.

The shape of the product of two container extensions is the product of their
shapes. The product of such container extensions should allow us to get either
the left, or the right element in the product. Therefore, the positions of this
container extension is the sum of the positions of the two container
extensions. What is a valid position in a given shape? If the position is
\mintinline{Coq}{inl p} (or \mintinline{Coq}{inr p}), then it is only valid if
\mintinline{Coq}{p} is valid in the first (resp. second) component of the
shape. The product of our two container extensions is
\mintinline{Coq}{Prod Id Id}.

Finally, the shape the sum of two container extensions is the sum of their
shapes. The positions are, like for products, the sum of their shapes. But to
check the validity of shapes in positions, their constructor must match, e.g.
shape \mintinline{Coq}{inl s} and shape \mintinline{Coq}{inl p}, with
\mintinline{Coq}{valid s p}.

Our final container extension is the following: 
\mintinline{Coq}{Sum (Const nat) (Prod Id Id)}.
The shape of this functor is: 
\mintinline{Coq}{nat + unit * unit}. The positions of this functor
are \mintinline[breaklines]{Coq}{False + (unit + unit)}, which is isomorphic
to \mintinline{Coq}{unit + unit}, which will only be valid in the RHS.

The Coq value \mintinline{Coq}{inl n} of type
\mintinline{Coq}{nat + bool * bool} would look as follows as a container
extension 
\mintinline{Coq}{App (Sum (Const nat) (Prod Id Id)) bool}:
\begin{minted}{Coq}
{| shape := inl n; cont := fun e => uninhabited e|}
\end{minted}
We omit the definition of \mintinline{Coq}{uninhabited}, which exploits the
fact that \mintinline{Coq}{e} for \mintinline{Coq}{inl n} implies
\mintinline{Coq}{False} to build an element of type \mintinline{Coq}{bool}.

Similarly, the Coq value \mintinline{Coq}{inr (true, false)} as a container
extension is as follows:
\begin{minted}{Coq}
  {| shape := inr (tt,tt);
     cont := fun e : {p | valid (inr (tt, tt)) p} =>
        match proj1_sig e with
        | inl f => match f with end
        | inr (inl tt) => fun _ => true
        | inr (inr tt) => fun _ => false
        end
  |}
\end{minted}
Here, the position \mintinline{Coq}{inl f} is impossible, as
\mintinline{Coq}{f} would need to be a proof of \mintinline{Coq}{False}.
Of the two remaining positions, the left one will return 
\mintinline{Coq}{true}, and the right position \mintinline{Coq}{false}.
\begin{figure}
\begin{minipage}{.52\textwidth}
  \begin{minted}[linenos, fontsize=\small]{Coq}
Context `(C1 : Cont S1 P1) 
        `(C2 : Cont S2 P2).

Definition Is := unit.
Definition Ip := unit.
Instance Id : Cont Is Ip := {
  valid := fun _ _ => true }.

Definition Ks (A : Set) := A.
Definition Kp := False.
Instance Const A : Cont (Ks A) Kp := {
  valid := fun _ p =>
    match p with end }.

Definition Ps : Type := S1 * S2.
Definition Pp : Type := P1 + P2.
Instance Prod : Cont Ps Pp := {
  valid := fun s p => 
    match p with
    | inl p' => valid (fst s) p'
    | inr p' => valid (snd s) p'
    end }.
\end{minted}
\end{minipage}
\begin{minipage}{.47\textwidth}
  \begin{minted}[linenos, firstnumber=23, fontsize=\small]{Coq}
Definition Ss : Type := S1 + S2.
Definition Sp : Type := P1 + P2.
Instance Sum : Cont Ss Sp := {
  valid := fun s p => 
    match s, p with
    | inl s', inl p' => valid s' p'
    | inr s', inr p' => valid s' p'
    | _, _ => false
    end }.

Definition Cs : Type := App F S2.
Definition Cp : Type := P1 * P2.
Instance Comp : Cont Cs Cp := 
{ valid :=  
    fun (s : App F S2) (p : P1 * P2) =>
        match valid (shape s) (fst p) 
        with
        | true  => fun H =>
            valid (cont s {fst p | H}) 
                  (snd p)
        | false => fun _ => false
        end eq_refl }.
  \end{minted}
\end{minipage}
  \caption{Polynomials as Container Extensions. In order: identity (L4-7),
  constant (L9-13), product (L15-22), sum (L23-31), composition (L33-44).}
  \label{fig:poly}
\end{figure}

@@

@@

\begin{minted}{Coq}
  Inductive AppR `{F : Cont Sh P} (X : Type) {e : equiv X}
           (x y : App F X) : Prop :=
  | AppR_ext
      (Es : shape x =e shape y)
      (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 =e cont y e2).
\end{minted}
The reason for this is to avoid a circularity in the definitions: arbitrary
equivalences for positions would force us to define the contents as morphisms
from positions to the contained elements, but that would as well force us to
redefine container extensions so that they are only applicable to types with
arbitrary equalities and morphisms.  We will see in the following section why
this is problematic for defining initial algebras of container extensions.
\dcas{I am actually not sure of what I wrote above, and it may be something I
simply overlooked. I will look into the Coq code and try to generalise it if I
have the time}

@@

  Corollary cata_ana_hylo `(F : Cont Sh P) `{equiv A} `{equiv B}
    (g : Alg F B) (h : RCoalg F A)
  : cata g \o rana h =e hylo g h.
\end{minted}
